---
title: "Flood susceptibility with GLO-30 HAND"
output: rmarkdown::html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

HAND (Height Above the Nearest Drainage) measures, for every cell of a terrain
model, how high that cell sits above the drainage channel it flows into. It is
a relative elevation, not an absolute one: a cell 2 m above its channel is
close to the water level whether it stands at 10 m or at 1000 m of altitude.
That makes it one of the cheapest proxies for flood susceptibility available
globally — low HAND means the terrain is near the level the water reaches.

`read_hand()` returns the GLO-30 HAND raster, derived from the Copernicus
30 m DEM, clipped to any polygon you give it.

```{r setup}
library(geohazards)
library(terra)
library(sf)
```

## Reading a raster for an area

`read_hand()` takes an `sf` object with POLYGON or MULTIPOLYGON geometry. Any
polygon works — a watershed, a neighbourhood, a study area drawn by hand — but
municipal boundaries are the most common case, and **geobr** provides the
official IBGE ones:

```{r read}
# Lauro de Freitas (BA): small, coastal, and covered by a single tile
muni <- geobr::read_municipality(code_muni = 2919207, year = 2022,
                                 simplified = FALSE, showProgress = FALSE)

hand <- read_hand(muni)

hand
plot(hand)
```

Nothing is downloaded in full. The GLO-30 HAND tiles are Cloud Optimized
GeoTIFFs, so only the blocks that intersect the polygon travel over the
network. `read_hand()` reports how many 1×1 degree tiles the area spans, which
is a good sanity check: a request covering many tiles is a large request, even
with COG.

Values are in **metres above the nearest drainage**, and the raster is masked
to the boundary, so cells outside the polygon are `NA`.

### Coordinate systems

The dataset is published in WGS84 (EPSG:4326), and that is what comes back by
default, whatever the CRS of `place` — the polygon is reprojected internally
for tile discovery, never the other way round.

Degrees are awkward for anything involving area or distance. Pass an
appropriate projected CRS to get a raster in metres:

```{r crs}
# UTM zone 24S, metric units
hand_utm <- read_hand(muni, crs_output = 31984)

res(hand_utm)  # cell size in metres
```

Reprojection uses bilinear resampling, which is the right choice for a
continuous surface. Do it once, at reading time, rather than reprojecting
derived products later.

### Several polygons at once

When `place` has more than one feature, the features are unioned and the
raster is masked to the combined boundary — one raster, not a list:

```{r multiple}
# Lauro de Freitas and its neighbour Simoes Filho
neighbours <- geobr::read_municipality(code_muni = c(2919207, 2930709),
                                       year = 2022, showProgress = FALSE)
hand_pair <- read_hand(neighbours)
```

## Reading the values

The interpretation is ordinal: lower means more susceptible. There is no
universal cut-off that turns HAND into a flood map, and the sensible threshold
depends on the basin, the channel network used to derive the product and the
return period you have in mind. Values in the single digits of metres are the
usual starting point, and the result should be treated as a screening layer,
not as a flood hazard map.

```{r distribution}
# Where the values sit for this area
global(hand, fun = "mean", na.rm = TRUE)
quantile(values(hand, na.rm = TRUE), c(0.05, 0.25, 0.5, 0.75, 0.95))

hist(values(hand, na.rm = TRUE), breaks = 50,
     main = "HAND distribution", xlab = "Metres above nearest drainage")
```

A screening mask is a single comparison:

```{r mask}
lowland <- hand < 5
plot(lowland, main = "Terrain within 5 m of the nearest drainage")

# Share of the municipality below the threshold
global(lowland, fun = "mean", na.rm = TRUE)
```

Classifying into bands is often more honest than a binary mask, because it
keeps the gradient visible:

```{r classify}
bands <- matrix(c(
   0,  2, 1,   # very low terrain, closest to the channel
   2,  5, 2,
   5, 15, 3,
  15, Inf, 4   # well above the drainage
), ncol = 3, byrow = TRUE)

hand_class <- classify(hand, bands)
plot(hand_class, main = "HAND bands")
```

## Extracting values for other features

The most useful thing about a HAND raster is rarely the raster itself: it is
the value under the features you already care about — buildings, schools,
census tracts, risk sectors.

```{r extract}
sectors <- read_sgb("risk", municipality = "Angra dos Reis", state = "RJ")
hand_angra <- read_hand(sectors)

sectors$hand_mean <- extract(
  hand_angra, vect(st_transform(sectors, 4326)), fun = mean, na.rm = TRUE
)[, 2]

sectors$hand_min <- extract(
  hand_angra, vect(st_transform(sectors, 4326)), fun = min, na.rm = TRUE
)[, 2]
```

Sectors that the SGB mapped as high geological risk *and* that sit a couple of
metres above the nearest drainage are where two independent lines of evidence
agree. See the article on the SGB cartography for what those sectors contain.

For point data, the same call returns one row per point:

```{r extract-points}
points <- st_as_sf(data.frame(lon = c(-44.32, -44.30), lat = c(-22.97, -22.99)),
                   coords = c("lon", "lat"), crs = 4326)
extract(hand_angra, vect(points))
```

## Mapping

The raster is a plain `SpatRaster`, so every terra and ggplot2 workflow
applies. A palette that goes from wet to dry reads better than the default:

```{r map}
plot(hand, col = hcl.colors(50, "Blues", rev = TRUE),
     main = "Height above nearest drainage (m)")
plot(st_geometry(muni), add = TRUE, border = "grey30")
```

With **ggplot2**, convert to a data frame first:

```{r ggplot}
library(ggplot2)

hand_df <- as.data.frame(hand, xy = TRUE)
names(hand_df)[3] <- "hand"

ggplot(hand_df, aes(x, y, fill = hand)) +
  geom_raster() +
  scale_fill_viridis_c(name = "HAND (m)", direction = -1) +
  coord_sf(crs = 4326) +
  theme_minimal()
```

## Exporting

```{r export}
writeRaster(hand, "hand.tif", overwrite = TRUE)

# Smaller file, same values, for sharing
writeRaster(hand, "hand_compressed.tif", overwrite = TRUE,
            gdal = c("COMPRESS=DEFLATE", "PREDICTOR=3"))
```

## Caveats

- **HAND is not a flood map.** It describes terrain, not water. It knows
  nothing about rainfall, river discharge, drainage infrastructure, tides or
  dams, and a cell can be low above its channel and still rarely flood.
- **The drainage network is modelled**, derived from the DEM, so HAND inherits
  every error of the underlying elevation model. Dense urban areas and flat
  floodplains are where it struggles most.
- **30 m cells** do not resolve individual buildings or street-level drainage.
- **Coastal flooding** driven by the sea, rather than by a channel, is outside
  what HAND represents.

## Source

GLO-30 HAND, published on the
[AWS Open Data registry](https://registry.opendata.aws/glo-30-hand/) and
derived from the Copernicus GLO-30 DEM. The HAND terrain descriptor was
introduced by [Nobre et al.
(2011)](https://www.sciencedirect.com/science/article/abs/pii/S0022169411002599).
Municipality boundaries come from the IBGE mesh, through **geobr**.
