---
title: "Mapping geological risk with the SGB"
output: rmarkdown::html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

The Geological Survey of Brazil (SGB/CPRM) maps, municipality by municipality,
the sectors where people live under geological risk — landslides, debris flows,
flash floods. `geohazards` reads that cartography directly into R.

This article goes through a full workflow: finding out what exists, downloading
it, summarising the exposure, mapping it interactively and exporting it.

```{r setup}
library(geohazards)
```

## What exists before you download it

The SGB mapping is not universal, and the server is slow enough that it pays to
ask before downloading. `sgb_inventory()` counts features without transferring
any geometry.

```{r inventory}
# The catalogue of products and the identifier each one is requested by
sgb_products()

# What the SGB holds for one municipality
sgb_inventory("Angra dos Reis", state = "RJ")

# ... or for a whole state
sgb_inventory(state = "RJ", by = "state")
```

If a municipality name is ambiguous, `state` disambiguates it. The IBGE code is
never ambiguous, and `sgb_municipalities()` finds it offline:

```{r lookup}
sgb_municipalities(search = "angra")
```

## Reading the risk sectorisation

```{r read}
angra <- read_sgb_risk("Angra dos Reis", state = "RJ")

nrow(angra)
names(angra)
sf::st_crs(angra)$epsg  # 4674, SIRGAS 2000
```

Field names come back in English, while the values remain as the SGB publishes
them — `risk_level` takes `"Alto"` and `"Muito alto"`. The mapping from the
original SGB field names is documented in `?read_sgb`.

## Exposure in risk sectors

`n_people`, `n_buildings` and `n_households` are the SGB's own per-sector
estimates. Summarising them takes one line, so the package does not wrap it:

```{r exposure}
exposure <- sf::st_drop_geometry(angra)

# Total exposure
colSums(exposure[c("n_people", "n_buildings", "n_households")], na.rm = TRUE)

# Broken down by risk level
aggregate(
  cbind(n_people, n_buildings, n_households) ~ risk_level,
  data = exposure, FUN = sum, na.rm = TRUE
)
```

The same works per process type, which is often the more actionable cut:

```{r exposure-by-process}
aggregate(n_people ~ process_type + risk_level, data = exposure,
          FUN = sum, na.rm = TRUE)
```

## An interactive map

`geohazards` does not depend on a mapping package: the sectors are a plain `sf`
object, so any of them will do. With **mapview**, colouring by risk level over
satellite imagery takes a few lines.

```{r map-simple}
library(mapview)

mapview(
  angra,
  zcol = "risk_level",
  col.regions = c("Alto" = "#F59E0B", "Muito alto" = "#DC2626"),
  map.types = c("Esri.WorldImagery", "OpenStreetMap", "CartoDB.Positron"),
  layer.name = "Risk sectors"
)
```

### A readable popup

The raw layer carries a few dozen fields, many of them administrative metadata,
and the occurrence layers carry an `email` field taken from the original
report — personal data that has no place in a popup. Selecting the fields that
matter and labelling them makes the map far easier to read:

```{r map-popup}
library(leafpop)

fields <- c("locality", "process_type", "process_subtype", "risk_level",
            "survey_date", "n_people")
labels <- c("Locality", "Process", "Specific type", "Risk level",
            "Surveyed", "People")

popup_data <- angra[, fields]
names(popup_data)[match(fields, names(popup_data))] <- labels

mapview(
  angra,
  zcol = "risk_level",
  col.regions = c("Alto" = "#F59E0B", "Muito alto" = "#DC2626"),
  map.types = "Esri.WorldImagery",
  layer.name = "Risk sectors",
  popup = popupTable(popup_data, feature.id = FALSE, row.numbers = FALSE)
)
```

To colour by a combination of fields — say risk level *and* process type —
build the combined category first:

```{r map-combined}
angra$risk_process <- paste(angra$risk_level, angra$process_type, sep = " | ")
mapview(angra, zcol = "risk_process", map.types = "Esri.WorldImagery")
```

## Exporting

Again, no wrapper needed: `sf::st_write()` handles every format.

```{r export}
# GeoPackage preserves long field names and accents
sf::st_write(angra, "angra_risk.gpkg", delete_dsn = TRUE)

# KML expects WGS84
sf::st_write(sf::st_transform(angra, 4326), "angra_risk.kml", delete_dsn = TRUE)
```

Avoid shapefile if you can: it truncates field names to ten characters, which
turns `n_affected_buildings` into something unreadable.

## Combining with flood susceptibility

The risk sectors say where people are exposed; `read_hand()` says how high the
terrain sits above the nearest drainage, which is a proxy for flood
susceptibility. The two combine directly:

```{r hand}
boundary <- geobr::read_municipality(code_muni = 3300100, year = 2022,
                                     showProgress = FALSE)
hand <- read_hand(boundary)

# Mean HAND under each risk sector
angra$hand_mean <- terra::extract(
  hand, terra::vect(sf::st_transform(angra, 4326)), fun = mean, na.rm = TRUE
)[, 2]
```

Sectors with a low mean HAND and a high risk level are the ones where flooding
and mapped geological risk coincide.

## Technical reports

The sectorisation of a municipality is usually documented in a technical
report, deposited in RIGeo, the SGB's institutional repository. Given its
handle, `read_sgb_report()` downloads and extracts the whole package:

```{r rigeo}
report <- read_sgb_report("doc/17701")

report$item$title
report$files
```

## Source

Geological Survey of Brazil (SGB/CPRM) — GeoSGB and RIGeo. The data is public;
please cite the source when using it. Municipality boundaries and codes come
from the IBGE mesh.
